A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-08 latex/main.tex
Sally's updates

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.

@@

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of multiple species~\cite{Felsenstein2004} using small numbers of widely diverged
genomes. In this context, recombination is a nuisance 
requiring careful consideration~\cite{Rannala2025recombination}
as it affects the inference of species trees~\cite{Posada2002effect}
and biases downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of genomes sampled from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods % citation for this?
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modelling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly limited to
visualisation, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

The global
surveillance data collected during the COVID-19 pandemic
presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer amount of data 
overwhelmed even approximate distance matrix-based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February, 2022 $\sim$80,000 samples per day
were submitted to GISAID~\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October 2025.
Regularly re-inferring trees de novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real time~\cite{McBroome2021,Kramer2023}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is an important force
in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present
in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{Varabyou2021rapid,%Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
As the range of applications for ARGs has broadened, the term is no
longer synonymous with the coalescent with 
recombination population genetic model \cite{Hudson1983,Griffiths1997ancestral}
and an ARG is now more usually regarded as a data structure
for representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
Strong software support for ARGs has facilitated these developments,
and the tskit library~\cite{Ralph2020} is the basis of a 
rapidly growing software ecosystem of
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that sequentially adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that the non-recombinant portions of this ARG are congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and that the ARG accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated through a community effort, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

>>>>

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes, 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.

@@

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of divergent species~\cite{Felsenstein2004}. Recombination among lineages, however, distorts the inference of species trees~\cite{Posada2002effect},
 biases downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}, and so requires careful consideration~\cite{Rannala2025recombination}.
Modern pathogen datasets are not sparsely sampled from different species
% Need to concrete. "large" means 100 to some people.
but consist of tens to hundreds of thousands of genomes sampled from 
% JK: deleted point about recomb here as TB, eg, doesn't
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to infer species trees~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of pathogen datasets are typically built with
simpler distance methods % citation for this?
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly restricted to
visualisation, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

The global
surveillance data collected during the COVID-19 pandemic
presents fundamental challenges to existing approaches. 
First, the sheer amount of data 
overwhelmed even approximate distance matrix-based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February 2022, $\sim$80,000 samples per day
were submitted to GISAID~\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October 2025.
Regularly re-inferring trees de novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real time~\cite{McBroome2021,Kramer2023}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
Recurrent recombination has posed an additional challenge to reconstructing the evolutionary history of
SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events impacting both the genealogy and disease attributes of the virus \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{Varabyou2021rapid,%Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated in a post hoc manner, after phylogenetic reconstruction.

Ancestral recombination graphs (ARGs) describe the reticulate 
genealogies of sampled sequences in a recombining species\cite{Wong2024},
and provide a natural and efficient means of 
systematically incorporating recombination 
into the study of SARS-CoV-2.
While ARGs have been of theoretical interest in population genetics 
for several decades\cite{Hudson1983,Griffiths1991two,Griffiths1997ancestral},
recent breakthroughs in
simulation\cite{Kelleher2016,Kelleher2018efficient,Haller2019,Gopalan2025bridging}
and inference~\citep{Rasmussen2014,Speidel2019,Kelleher2019,
Wohns2022,Zhang2023,Gunnarsson2024,Deng2025robust} have led to a surge 
of interest in their application in population and statistical 
genetics\cite{Brandt2024,Lewanski2024,Nielsen2025}.
% This is necessary because a bunch of the phylo literature refers
% to ARGs, but it's all in the "coalescent with recombination" sense
% Rannala2025 dismisses ARG-based methods at a stroke because they're
% all based on the assumption of a single panmictic population 
As the range of applications for ARGs has broadened, the term has itself evolved from a specific model of coalescence with 
recombination  \cite{Hudson1983,Griffiths1997ancestral}to a rich data structure representing recombinant ancestry\cite{Minichiello2006mapping,Wong2024}.
The development of extensive software support has facilitated the broader application of ARGS. In particular, the collaboratively built tskit library~\cite{Ralph2020} underpins a 
rapidly growing software ecosystem of
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit,Talbot2025argscape},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that sequentially adds samples over time in a similar manner 
to UShER, while automatically detecting and incorporating recombination
in real-time. Based on the freely-available Viridian dataset, a curated
% NOTE: Viridian doesn't provide alignments, we do that with MAFFT
collection of SARS-CoV-2 genomes purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that the non-recombinant portions of this ARG are congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020}.
The ARG also accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated through a community effort and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic and to prepare for the next.
